\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-01-31 move_schedules.tex
new plots, new revision to 3.4.3, and small typo fixes

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract}
\abstract{
	The standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications.
%	As our first result, we propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that takes $O(m^2)$ preprocessing time uses only $O(m)$ space. 
%	We also show that this maintenance procedure yields a factor-2 net reduction in column operations over the state of the art approach. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that requires only $O(m \log \log m)$ time and $O(m)$ space. 
By reduction to a particular longest common subsequence problem, we show the storage needed to employ this strategy is actually sublinear in expectation. 
	Exploiting this connection, we show experimentally the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
	Applications to movie data, dynamic metric spaces, and multi-parameter persistence are also presented.
}
%\end{abstract}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
% With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices across thousands of snapshots in time.
Acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time via a decomposition $R = D V$ of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
In particular, the vineyards algorithm transforms a time-varying filtration into a particular set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams independently at every time point.

@@

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

\subsection{Related Work}\label{sec:related_work} 
To the authors knowledge, work focused on ways of updating a  decomposition $R = DV$, for all homological dimensions, is limited---there is the vineyards algorithm~\cite{cohen2006vines} and the moves algorithm~\cite{busaryev2010tracking}, both of which are discussed extensively in section~\ref{sec:background}.

In contrast, there is extensive work  on improving the efficiency of computing a (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explicitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

By making stronger assumptions on the underlying topological space, restricting the homological dimension, or targeting  a weaker invariant (e.g. Betti numbers), one can usually obtain  faster or alternative approaches  recording information related to persistence.
For example, Attali et al.~\cite{attali2009persistence} give a linear time algorithm (under the $\texttt{word RAM}$ model of computation) for computing persistence on graphs.
In the same paper, they describe how to obtain $\epsilon$-simplifications of $1$-dimensional persistence diagrams for filtered $2$-manifolds by using duality and symmetry theorems. 
Along a similar vein, Edelsbrunner et al.~\cite{edelsbrunner2000topological} give a fast incremental algorithm for computing persistent Betti numbers up to dimension $2$, again by utilizing symmetry, duality, and ``time-reversal''~\cite{delfinado1995incremental}. Chen et al.~\cite{chen2013output} give an output-sensitive method for computing persistent homology,  utilizing the property that certain submatrices of  $D$ have the same rank as  $R$, which they exploit through fast sub-cubic rank algorithms specialized for sparse-matrices.

If  zeroth homology is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give a $O(e)$-per-update algorithm that allows one to maintain a \emph{merge tree} with $e$ edges,
associated to the filtration parameter changing continuously over time. 
If only the zeroth-dimensional Betti numbers are needed for a particular application, this problem reduces even further to tracking the connected components of a dynamic graph---sometimes referred to as the \emph{dynamic connectivity problem}, which can be efficiently  solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

 \subsection{A Motivating Example}\label{sec:motivation} 
We illustrate why the vineyards algorithm does not always yield an efficient strategy for time-varying setting with a simple experiment. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid and its associated sublevel-set filtrations:   
\begin{center}
	\vspace{1em}
	\includegraphics[height=1.25in]{circle_complex.png}
\end{center}
Each image contains a fixed number of pixels whose intensities vary with time. For each image, we build a fixed-size filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane whose inclusion maps are obtained by lower stars of pixel values.   
%Snapshots of the image data and its corresponding time-varying filtration are shown below Figure~\ref{fig:vineyards}. 
%\begin{figure*}[h!]
%\centering	
%\end{figure*}
%Note that the size of the filtration is fixed---no simplices are added or removed over time. 
Two events alter the persistence diagrams over time: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus  splits into four components. 
From left to right, the Betti numbers of the five evenly spaced `snapshots' of the filtration shown above are: $(\beta_0,\beta_1)  = (1,0), \; (1,1),  \; (1,1), \; (1,1), \; (4,0)$.
Thus, in this example, only a few persistence diagrams are needed to capture the major changes to the topology. 
\begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.75\textwidth]{circle_vineyards_vector.pdf}
%	\includesvg{circle_vineyards}
 	\caption{The cumulative column operations needed to compute persistence across the time-varying filtration of grayscale images. Observe 10 independent persistence computations evenly spaced in time (green line) captures the major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}

  We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
  Since it is unknown a priori at when the persistent pairing function changes, one solution is to do $n$ independent persistence computations $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between contiguous time-points and decompose these homotopies into adjacent transpositions in the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, though it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family, which we'll refer to as the \emph{continuous setting}.

 The cumulative cost (in total column operations) of these various approaches are shown on the left side of Figure~\ref{fig:vineyards}, wherein the reduction algorithm (\texttt{pHcol}) and the vineyards algorithm are compared. Two approximate strategies (green and purple) and two continuous strategies (black and blue) are shown.
  Note that without knowing where persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
  In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when pairing function is unchanged. 
Indeed, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is more efficient than iteratively update the decomposition, as shown by the case where $n = 10$.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed.

\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of the number of valid intermediate decomposition states, as compared to vineyards---for dynamic persistence. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

\subsubsection{Proxy objective}\label{sec:proxy_objective}
Although the greedy approach from the last section can lead to arbitrarily high cost schedules, we find similar greedy-like strategies can yield computationally efficient heuristics in practice, even if they are suboptimal. 
Choosing the lowest-cost move takes $\approx O(dm)$---implying a greedy strategy would require $O(d^2 m)$ time---which is computationally viable as a heuristic when $d << m$. 
We seek a fast procedure for generating schedules that is not only substantially better than a randomly or naive strategy in term of column reductions, but is also low complexity enough that it is practical in a real-world setting.
For example, the vineyards algorithm requires $O(m \log m + k)$ preprocessing time to compute a schedule of size $k \sim O(m^2)$ transpositions, thus for Algorithm~\ref{alg:schedule} to be considered a more efficient alternative to vineyards, a solution subquadratic in $m$ is desirable. 
%In what follows, we present such an objective, showing its relation to the problem at hand as well its use in other known computational problems. 
% We show empirical results demonstrating its practical effectiveness as a heuristic in section~\ref{sec:results}.

@@

All three schedules were generated from the same $\mathrm{LCS}(p,q) = (4\,7\,8)$ and all three schedules transform $p \mapsto q$ in $d = 6$ moves. 

@@

The example above suggests minimizing~\eqref{eq:sorting_additive} is reminiscent of the $k$-layer crossing minimization problem for $k$ sets of permutations, which is NP-hard for $k \geq 4$~\cite{biedl2009complexity}. 

@@

% TODO: Move computational results to subsection, add applications
% \subsection{Empirical results}\label{sec:results}
% \section{Applications and Discussion}\label{sec:applications}
\section{Applications and Experiments}\label{sec:results}
\subsection{Movies}
A common application of persistence is to characterizing topological structure in image data. Since a  set of ``snapshot'' frame of a movie can be equivalently thought of as discrete 1-parameter family, our framework provides a natural extension of the typical image analysis to video data. 
To demonstrate the benefit of using minimal sized scheduled and the scalability of the greedy approach proposed in section~\ref{sec:proxy_objective}, we re-use the movie data from~\ref{sec:motivation} as a baseline benchmark. We perform two performance tests: one to  one to test the impact of shrinking the number of permutations a given reindexing operation needs and one to test the asymptotic behavior of the greedy approach and.

>>>>

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract}
\abstract{
	The standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices is the reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications.
%	As our first result, we propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that takes $O(m^2)$ preprocessing time uses only $O(m)$ space. 
%	We also show that this maintenance procedure yields a factor-2 net reduction in column operations over the state of the art approach. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that requires only $O(m \log \log m)$ time and $O(m)$ space to construct. 
By reduction to a particular longest common subsequence problem, we show the storage needed to employ this strategy is actually sublinear in expectation. 
Exploiting this connection, we show experimentally the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
}
%\end{abstract}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With state of the art tools, there is no recourse when faced with a time-varying complex containing millions of simplices across thousands of snapshots in time.
Moreover, acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes all immediately become computationally viable tools in dynamic settings. Persistence-based examples include adaptive template functions~\cite{polanco2019adaptive}, persistence images~\cite{adams2017persistence}, and $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time via a decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
In particular, the vineyards algorithm transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

@@

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

\subsection{Related Work}\label{sec:related_work} 
To the authors knowledge, work focused on ways of updating a  decomposition $R = DV$, for all homological dimensions, is limited: there is the vineyards algorithm~\cite{cohen2006vines} and the moves algorithm~\cite{busaryev2010tracking}, both of which are discussed extensively in section~\ref{sec:background}. At the time of writing, we were made aware of very recent work~\cite{luo2021accelerating} that iteratively repairs a permuted decomposition via a column swapping and reduce strategy, which they call ``warm starts.'' Though similar in spirit, their approach relies on the reduction algorithm as a subprocedure, which is quite different from the strategy we employ here.

Contrasting the dynamic setting, there is extensive work  on improving the efficiency of computing a single (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explicitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

By making stronger assumptions on the underlying topological space, restricting the homological dimension, or targeting  a weaker invariant (e.g. Betti numbers), one can usually obtain faster algorithms.
For example, Attali et al.~\cite{attali2009persistence} give a linear time algorithm for computing persistence on graphs.
In the same paper, they describe how to obtain $\epsilon$-simplifications of $1$-dimensional persistence diagrams for filtered $2$-manifolds by using duality and symmetry theorems. 
Along a similar vein, Edelsbrunner et al.~\cite{edelsbrunner2000topological} give a fast incremental algorithm for computing persistent Betti numbers up to dimension $2$, again by utilizing symmetry, duality, and ``time-reversal''~\cite{delfinado1995incremental}. Chen et al.~\cite{chen2013output} give an output-sensitive method for computing persistent homology,  utilizing the property that certain submatrices of  $D$ have the same rank as  $R$, which they exploit through fast sub-cubic rank algorithms specialized for sparse-matrices.

If  zeroth homology ($p = 0$) is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

 \subsection{A Motivating Example}\label{sec:motivation} 
We illustrate why the vineyards algorithm does not always yield an efficient strategy for time-varying setting with a simple experiment. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid and its associated sublevel-set filtrations:   
\begin{center}
	\vspace{1em}
	\includegraphics[height=1.25in]{circle_complex.png}
\end{center}
Each image contains a fixed number of pixels whose intensities vary with time. For each image, we build a fixed-size filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane whose inclusion maps are obtained by lower stars of pixel values.   
%Snapshots of the image data and its corresponding time-varying filtration are shown below Figure~\ref{fig:vineyards}. 
%\begin{figure*}[h!]
%\centering	
%\end{figure*}
%Note that the size of the filtration is fixed---no simplices are added or removed over time. 
Two events critically change the pairing function: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus  splits into four components. 
From left to right, the Betti numbers of the five evenly spaced `snapshots' of the filtration shown above are: $(\beta_0,\beta_1)  = (1,0), \; (1,1),  \; (1,1), \; (1,1), \; (4,0)$.
Thus, in this example, only a few persistence diagrams are needed to capture the major changes to the topology. 
\begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.75\textwidth]{circle_vineyards_vector.pdf}
%	\includesvg{circle_vineyards}
 	\caption{The cumulative column operations needed to compute persistence across the time-varying filtration of grayscale images. Observe 10 independent persistence computations evenly spaced in time (green line) captures the major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}

  We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
  Since it is unknown a priori at when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions of the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is only not guaranteed to capture all homological changes in persistence that occur in simulating a continuous 1-parameter family of diagrams.

 The cumulative cost (in total column operations) of these various approaches are shown on the left side of Figure~\ref{fig:vineyards}, wherein the reduction algorithm (\texttt{pHcol}) and the vineyards algorithm are compared. Two discrete strategies (green and purple) and two continuous strategies (black and blue) are shown.
  Note that without knowing where persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
  In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when pairing function is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of the number of valid intermediate decomposition states, as compared to vineyards---for dynamic persistence. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

\subsubsection{Proxy objective}\label{sec:proxy_objective}
Although the greedy approach from the last section can lead to arbitrarily high cost schedules, we find similar greedy-like strategies can yield computationally efficient heuristics in practice, even if they are suboptimal. 
%Choosing the lowest-cost move takes $\approx O(dm)$---implying a greedy strategy would require $O(d^2 m)$ time---which is computationally viable as a heuristic when $d << m$. 
We seek a fast procedure for generating schedules that is not only substantially better than a random or simple schedule strategy in term of column reductions, but has a low enough time and storage complexity to be practical for larger data sets.
%For example, the vineyards algorithm requires $O(m \log m + k)$ preprocessing time to compute a schedule of size $k \sim O(m^2)$ transpositions, thus for Algorithm~\ref{alg:schedule} to be considered a more efficient alternative to vineyards, a solution subquadratic in $m$ is desirable. 
%In what follows, we present such an objective, showing its relation to the problem at hand as well its use in other known computational problems. 
% We show empirical results demonstrating its practical effectiveness as a heuristic in section~\ref{sec:results}.

@@

All three schedules were generated from the same $\mathrm{LCS}(p,q) = (4\,7\,8)$ and each schedule transforms $p \mapsto q$ in $d = 6$ moves. 

@@

%The greedy heuristic proposed to minimize equation~\ref{eq:schedule_opt_succ} was used to produce $S_1$, which has 21 edge crossings. 

@@

% TODO: Move computational results to subsection, add applications
% \subsection{Empirical results}\label{sec:results}
% \section{Applications and Discussion}\label{sec:applications}
\section{Applications and Experiments}\label{sec:results}
\subsection{Video data}
A common application of persistence is to characterizing topological structure in image data. Since a  set of ``snapshot'' frames of a video can be equivalently thought of as discrete 1-parameter family, our framework provides a natural extension of the typical image analysis to video data. 
To demonstrate the benefit of using minimal sized scheduled and the scalability of the greedy approach proposed in section~\ref{sec:proxy_objective}, we re-use the video data from~\ref{sec:motivation} as a baseline benchmark. We perform two performance tests: one to  one to test the impact of shrinking the number of permutations a given reindexing operation needs and one to test the asymptotic behavior of the greedy approach and.

 In the first test, we fix a grid size of $9 \times 9$ and record the cumulative number of column operations needed to simulate persistence dynamically across 25 evenly-spaced time points using a variety of scheduling strategies.
 The primary three strategies we test are the greedy approach from section~\ref{sec:proxy_objective}, the ``simple'' approach which uses upwards of $O(m)$ move permutations via selection sort, and a third strategy which interpolates between the two. 
 In particular, if $d = m - \lvert \mathrm{LCS}(f, f') \rvert$ we use a parameter $\alpha \in [0,1]$ to choose $m - \alpha \cdot d$ random symbols to move using the same construction method outlined in section~\ref{sec:schedule_construction}. When $\alpha = 0$, the strategy reduces to using selection sort to reindex $f \mapsto f'$ using a random ordering of simplices; otherwise, $\alpha = 1$ reduces to using a minimal sized schedule (with a random ordering of simplices). The results are summarized in the left graph on Figure~\ref{fig:movie_perf}. To get an idea of the variation of the random sampling, we run 10 independent iterations and shade the upper and lower bounds of the schedule costs, which is shown around the mean schedule cost depicted by the solid lines in Figure~\ref{fig:movie_perf}. 
 As one can see from the figure, while using less move operations (lower $\alpha$) does progressively reduce column operations, constructing random schedules of minimal size is no more competitive than the selection sort strategy. This suggests that efficient schedule construction needs to account for the structure of performing several permutations in sequence, like the greedy heuristic we introduced, to yield a performance boost.
